sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-19 doc/paper.Rmd

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues, relatively flat likelihood surfaces, to impart prior information, or to achieve regularization.
Following other recent SPDE implementations [@simpson2017; @osgood-zimmerman2021], penalized complexity priors (`pc_matern`) may be particularly useful to constrain the spatial range and variance parameters.

@@

Take home, re-iterate paragraph?

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru. 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data and delta/hurdle models in VAST.
Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1`).
However, for users who require multivariate spatiotemporal models or the Poisson-link delta model (REF), there is no substitute for VAST.

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences will depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (Supp fig?).
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
We think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB, and sdmTMB works with regular R data frames whereas inlabru works with `sp` class data objects.
However, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv, but, yikes, do we have space:

- mgcv is capable of fitting with the SPDE (REF)
- typical smoothers don't offer MVN interpretability
- Tweedie is a big factor... must fix power 'p' to fit a GAMM.
- Barrier approaches are different
- doesn't have functionality built in for indexes/COG and uncertainty on such derived parameters
- for this class of models, we think sdmTMB has advantages, but mgcv is an immensely powerful modelling package that extends well beyond spatial/spatiotemporal data.

Final paragraph touching on other stuff possible in one paragraph: priors, Bayes, barrier meshes, randomized quantile residuals, simulation from joint precision, ...?

>>>>

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations [@simpson2017; @breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors (`pc_matern()`) may be particularly useful to constrain the spatial range and variance parameters.

@@

<!-- Take home, re-iterate paragraph? -->

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the two most closely related are VAST and inlabru. 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is a less modular interface (e.g., automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some of the features that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves 3 of the 4 design principles outlined in Thorson (REF) for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.

<!--         -   "VAST represents four main design principles. Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
<!--         -   Condition on missing covariates (check, random fields) -->
<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->

inlabru is similar to sdmTMB in design, but differs fundamentally in that fitting is achieved via INLA and approximate Bayesian computation.
For some models, TMB/sdmTMB can be considerably faster than INLA/inlabru [@osgood-zimmerman2021], but in our experience, those differences depend on many aspects of the data, SPDE mesh, model, and fitting algorithm settings (TODO Supp fig?).
One major capability included in sdmTMB but not INLA/inlabru is anisotropy.
Furthermore, we think the sdmTMB interface will be more familiar than inlabru to users of `glm()`, lme4, mgcv, or glmmTMB.
On the other hand, inlabru fits additional model classes including log Gaussian Cox processes and may be particularly usable to those already familiar with INLA.

mgcv can fit many similar models to sdmTMB either with various flavours of splines or even with the SPDE approximation to random fields (REF); however, there are some notable advantages of sdmTMB for the specific class of models discussed here.
First, the smoothers typically used with mgcv (thin plate splines, ...), while capable of making equivalent predictions (REF) do not offer the ecological interpretability of a spatially explicit random field model (spatial range and variance).
Second, the implementation of the Tweedie distribution in mgcv requires the power parameter to be fixed in a mixed effects model.
Third, mgcv does not have built-in functionality to calculate derived values with uncertainty such as population indexes or center of gravity.
Despite these sdmTMB advantages for the models discussed here, mgcv is an immensely powerful modelling package that extends well beyond spatial random field models.

Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes (REF) and improve ecological management (REF).
These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
We hope the development of user-friendly interfaces such as sdmTMB opens this useful class of models to a wider audience of users.
